\chapter{习题 19: 一个简单的对象系统}

我是先学了 C 语言后才学习的面向对象编程，而学习面向对象编程用的就是 C 语言，通过用 C 语言建立 
OOP 系统让我学到了 OOP 的基本概念。而你可能是先学了某种面向对象编程语言后才来学习 C 语言的，
所以这种学习路径对你应该也有帮助。本节习题中，你将要创建一个简单的对象系统，不过你还会学到关于 C 
预处理器（C Pre-Processor，CPP）的更多知识。

本章习题的内容是创建一个简单的游戏，你在游戏中的任务是在一座小城堡里杀死一只牛头怪。这个游戏
没什么花哨的地方，只不过是四间屋子和一个坏人而已。这个项目同时也是一个多文件项目，和你之前做过
的相比更类似于一个真正的 C 软件。这也是为什么我要在这里向你介绍 CPP，因为你在写多个文件的程序
时会用到它。

\section{CPP 的工作原理}

C Pre-Processor 是一个模板预处理系统。它的目的很明确，就是让 C 语言更易用，不过它的实现方式
是使用了一种语法敏感的模板机制。传统习惯上人们只把 CPP 用于存储常量以及创建用来简化重复工作的
“宏”。在现代 C 语言中你将使用 CPP 作为一个代码生成器，用来创建模板化的代码片段。

CPP 的工作模式是这样的：你提供一个文件，通常是一个 .c 文件，它会对其中以 \verb|#| 字符开头的
部分字串进行处理。对于每一处 \verb|#| ，它会对其中的文本输入进行一个特殊的替换处理。而它最
主要的优势是它可以 \emph{include} 其它文件，然后把该文件的内容扩充到前者中去。

快速检验 CPP 功能的方法就是将上一习题拿来运行如下命令：

\begin{Verbatim}
cpp ex18.c | less
\end{Verbatim}

你会收到很长的一串输出，不过逐页浏览下去你会看到你用 \verb|#include| 包含进来的其它文件
的内容。向下滚动到原始代码，你可以看到 \program{cpp} 是怎样依据头文件中的各种 \verb|#define| 
宏语句来修改源代码文件的。

C 编译器和 \program{cpp} 结合得非常紧密，它会直接运行 cpp，并且熟悉 cpp 的功能。在
现代 C 语言中，\program{cpp} 系统已经深入集成到 C 的功能中，你甚至可以直接认为它就是 C 
语言的一部分了。

在剩下的小节中，我将用到更多的 CPP 语法，并在使用过程中逐一向你解释。


\section{对象系统的原型}

我们要创建的 OPP 系统是一个简单的“原型(prototype)”式的类似于 JavaScript 对象系统的东西。
我们使用的不是类，取而代之的是你从包含属性栏位的 prototype 出发，然后使用这个 prototype 系统
创建其它的对象实例。和传统的基于 class 的设计相比，这种“无 class”的设计方式更容易实现和应对。

\subsection{对象头文件}

我要将数据类型和函数声明放到一个独立的称作 \file{object.h} 的头文件中。这是一个标准的 C 
语言实践模式，通过这种方式，你可以直接发布二进制库，不过还可以让程序员使用它进行编译。在这个
文件里我用了几个高级 CPP 技巧，我会快速对它们进行讲解，后面你可以看到它们在实际运行中的功用：

\begin{code}{object.h}
<< d['code/object.h|pyg|l'] >>
\end{code}

看一下这个文件，你可以看到我们用了几个之前没见过的语法：

\begin{description}

\item[\#ifndef] You've seen a \verb|#define| for making simple constants, but
    the CPP can also do logic and remove sections of code.  This \verb|#ifndef| is
    "if not defined" and checks if there's already a \verb|#define _object_h| and
    if there is it skips all of this code.  I do this so that we can include
    this file any time we want and not worry about it defining things multiple
    times.
\item[\#define] With the above \verb|#ifndef| shielding this file from we then
    add the \ident{\_object\_h} define so that any attempts to include it later
    cause the above to skip.
\item[\#define NEW(T,N)]  This makes a macro, and it works like a template function
    that spits out the code on the right, whenever you write use the macro
    on the left.  This one is simply making a short version of the normal way
    we'll call \func{Object\_new} and avoids potential errors with calling it
    wrong.  The way the macro works is the \ident{T} and \ident{N} parameters
    to \ident{NEW} are "injected" into the line of code on the right.  The
    syntax \verb|T##Proto| says to "concat Proto at the end of T", so if you had
    \verb|NEW(Room, "Hello.")| then it'd make \ident{RoomProto} there.
\item[\#define \_(N)] This macro is a bit of "syntactic sugar" for the object
    system and basically helps you write \verb|obj->proto.blah| as simply
    \verb|obj->_(blah)|.  It's not necessary, but it's a fun little trick
    that I'll use later.
\end{description}

\subsection{The Object Source File}

The \file{object.h} file is declaring functions and data types that are defined
(created) in the \file{object.c}, so that's next:

\begin{code}{object.c}
<< d['code/object.c|pyg|l'] >>
\end{code}

There's really nothing new in this file, except one \emph{tiny} little trick.
The function \func{Object\_new} uses an aspect of how \ident{struct}s work
by putting the base prototype at the beginning of the struct.  When you look
at the \file{ex19.h} header later, you'll see how I make the first field
in the struct an \ident{Object}.  Since C puts the fields in a struct
in order, and since a pointer just points at a chunk of memory, I can
"cast" a pointer to anything I want.  In this case, even though I'm taking
a potentially larger block of memory from \func{calloc}, I'm using a
\ident{Object} pointer to work with it.

I explain this a bit better when we write the \file{ex19.h} file since it's
easier to understand when you see it being used.

That creates your base object system, but you'll need a way to compile it
and link it into your \file{ex19.c} file to create a complete program.  The
\file{object.c} file on its own doesn't have a \func{main} so it isn't
enough to make a full program.  Here's a \file{Makefile} that will do this
based on the one you've been using:

\begin{code}{The Makefile}
\begin{lstlisting}
<< d['code/ex19.1.Makefile'] >>
\end{lstlisting}
\end{code}

This \file{Makefile} is doing nothing more than saying that \program{ex19}
depends on \ident{object.o}.  Remember how \program{make} knows how to build
different kinds of files by their extensions?  Doing this tells make the
following:

\begin{enumerate}
\item When I say run \program{make} the default \ident{all} should just build
    \program{ex19}.
\item When you build \program{ex19}, you need to also build \file{object.o}
    and include it in the build.
\item \program{make} can't see anything in the file for \file{object.o}, but it does
    see an \file{object.c} file, and it knows how to turn a \file{.c} into
    a \file{.o}, so it does that.
\item Once it has \file{object.o} built it then runs the correct compile
    command to build \program{ex19} from \file{ex19.c} and \file{object.o}.
\end{enumerate}


\section{The Game Implementation}

Once you have those files you just need to implement the actual game
using the object system, and first step is putting all the data types
and function declarations in a \file{ex19.h} file:

\begin{code}{ex19.h}
<< d['code/ex19.h|pyg|l'] >>
\end{code}

That sets up three new Objects you'll be using: \ident{Monster}, \ident{Room},
and \ident{Map}.

Taking a look at \file{object.c:52} you can see where I use a pointer
\verb|Object *el = calloc(1, size)|.  Go back and look at the \func{NEW}
macro in \file{object.h} and you can see that it is getting the \func{sizeof}
another struct, say \ident{Room}, and I allocate that much.  However, because
I've pointed a \ident{Object} pointer at this block of memory, and because
I put an \ident{Object proto} field at the from of \ident{Room}, I'm able
to treat a \ident{Room} like it's an \ident{Object}.

The way to break this down is like so:

\begin{enumerate}
\item I call \verb|NEW(Room, "Hello.")| which the CPP expands as a macro
    into \verb|Object_new(sizeof(Room), RoomProto, "Hello.")|.
\item This runs, and inside \func{Object\_new} I allocate a piece of memory
    that's \ident{Room} in size, \emph{but} point a \ident{Object *el} pointer
    at it.
\item Since C puts the \ident{Room.proto} field first, that means the \ident{el}
    pointer is really only pointing at enough of the block of memory to
    see a full \ident{Object} struct.  It has no idea that it's even called
    \ident{proto}.
\item It then uses this \ident{Object *el} pointer to set the contents of
    the piece of memory correctly with \verb|*el = proto;|.  Remember that
    you can copy structs, and that \ident{*el} means "the value of whatever el points
    at", so this means "assign the proto struct to whatever el points at".
\item Now that this mystery struct is filled in with the right data from
    \ident{proto}, the function can then call \ident{init} or \ident{destroy}
    on the \ident{Object}, but the cool part is whoever called this function
    can \emph{change} these out for whatever ones they want.
\end{enumerate}

And with that, we have a way to get this one function to construct new types,
and give them new functions to change their behavior.  This may seem like
"hackery" but it's stock C and totally valid.  In fact there's quite a few
standard system functions that work this same way, and we'll be using some of
them for converting addresses in network code.

With the function definitions and data structures written out I can now
actually implement the game with four rooms and a minotaur to beat up:

\begin{code}{ex19.c}
<< d['code/ex19.c|pyg|l'] >>
\end{code}

Honestly there isn't much in this that you haven't seen, and only you might
need to understand how I'm using the macros I made from the headers files.
Here's the important key things to study and understand:

\begin{enumerate}
\item Implementing a prototype involves creating its version of the
    functions, and then creating a single struct ending in "Proto".
    Look at \ident{MonsterProto}, \ident{RoomProto} and \ident{MapProto}.
\item Because of how \func{Object\_new} is implemented, if you don't set
    a function in your prototype, then it will get the default implementation
    created in \file{object.c}.
\item In \func{Map\_init} I create the little world, but more importantly
    I use the \ident{NEW} macro from \file{object.h} to build all of the
    objects.  To get this concept in your head, try replacing the \ident{NEW}
    usage with direct \ident{Object\_new} calls to see how it's being
    translated.
\item Working with these objects involves calling functions on them, and the
    \ident{\_(N)} macro does this for me.  If you look at the code
    \verb|monster->_(attack)(monster, damage)| you see that I'm using the
    macro, which gets replaced with \verb|monster->proto.attack(monster, damage)|.
    Study this transformation again by rewriting these calls back to their 
    original.  Also, if you get stuck then run \program{cpp} manually to see what it's
    going to do.
\item I'm using two new functions \func{srand} and \func{rand}, which setup
    a simple random number generator good enough for the game.  I also use
    \func{time} to initialize the random number generator.  Research those.
\item I use a new function \func{getchar} that gets a single character
    from the stdin.  Research it.
\end{enumerate}


\section{What You Should See}

Here's me playing my own game:

\begin{code}{ex19 output}
\begin{lstlisting}
<< d['code/ex19.out'] >>
\end{lstlisting}
\end{code}


\section{Auditing The Game}

As an exercise for you I have left out all of the \func{assert} checks
I normally put into a piece of software.  You've seen me use \func{assert} to
make sure a program is running correctly, but now I want you to go back
and do the following:

\begin{enumerate}
\item Look at each function you've defined, one file at a time.
\item At the top of each function, add \func{asserts} that make sure
    the input parameters are correct.  For example, in \func{Object\_new}
    you want a \verb|assert(description != NULL)|.
\item Go through each line of the function, and find any functions
    being called.  Read the documentation (man page) for that function,
    and confirm what it returns for an error.  Add another assert to
    check that the error didn't happen.  For example, in \func{Object\_new}
    you need one after the call to \func{calloc} that does \verb|assert(el != NULL)|.
\item If a function is expected to return a value, either make sure it returns
    an error value (like NULL), or have an assert to make sure that the returned
    variable isn't invalid.  For example, in \func{Object\_new}, you need
    to have \verb|assert(el != NULL)| again before the last return since
    that part can never be NULL.
\item For every \ident{if-statement} you write, make sure there's an else
    clause unless that if is an error check that causes an exit.
\item For every \ident{switch-statement} you write, make sure that there's
    a \ident{default} case that handles anything you didn't anticipate.
\end{enumerate}

Take your time going through every line of the function and find any errors you
make.  Remember that the point of this exercise is to stop being a "coder" and
switch your brain into being a "hacker".  Try to see how you could break it,
then write code to prevent it or abort early if you can.

\section{Extra Credit}

\begin{enumerate}
\item Update the \file{Makefile} so that when you do \program{make clean} 
     it will also remove the \file{object.o} file.
\item Write a test script that works the game in different ways and augment
    the \file{Makefile} so you can run \program{make test} and it'll thrash
    the game with your script.
\item Add more rooms and monsters to the game.
\item Put the game mechanics into a third file, compile it to .o, and then
    use that to write another little game.  If you're doing it right
    you should only have a new \ident{Map} and a \func{main} function
    in the new game.
\end{enumerate}

